Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-24 latex/main.tex
Updates on pango lineages from yan and shing

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
of interest in their application in population and statical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
A subtle but important point distinguishing recent work on ARGs
from the classical literature is that the ARG 
``data structure''\cite{Minichiello2006mapping,Wong2024} is no longer
tightly coupled to the generative
process\cite{Hudson1983,Griffiths1997ancestral}, substantially 
broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
Another important point is that strong software support for ARGs now
exists, and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing 
software ecosystem of 
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

>>>>

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2024robust} have led to a surge 
of interest in their application in population and statical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
A subtle but important point distinguishing recent work on ARGs
from the classical literature is that the ARG 
``data structure''\cite{Minichiello2006mapping,Wong2024} is no longer
tightly coupled to the generative
process\cite{Hudson1983,Griffiths1997ancestral}, substantially 
broadening the range of potential applications~\cite{Bisschop2025likelihoods}.
Another important point is that strong software support for ARGs now
exists, and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing 
software ecosystem of 
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.
